Likelihoods for a general class of ARGs under the SMC

### 2025-04-23 review-responses.tex

\begin{point}{\revref} % \&75
 Isn't it possible to run SINGER on up to thousands of samples? I know that several hundreds is also a thousand!
\end{point}

\reply{
}

\begin{point}{}
Mention that the standard SMC formulation was left-to-right
\end{point}

\reply{
We inserted a parenthetical comment stating that the SMC
is ``conventionally regarded as an along-the-genome process''.
}

### 2025-04-23 paper.tex
Address some comments

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. First,
the Bayesian inference of ARGs under the Sequentially Markov
Coalescent (SMC), is now practical for tens-to-hundreds of samples.
Second, approximate models and heuristics can now scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they
estimate do not have the topological properties required to compute a
likelihood under models such as the SMC under present-day formulations.
In particular, heuristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a
backwards-time formulation of the SMC
and derive a straightforward definition
of the likelihood of a general class of ARG under this model.
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies.
We discuss the possibilities for inference that this opens. \revpoint{2}{3}
\end{abstract}

>>>>

\begin{abstract}
Ancestral recombination graphs (ARGs) are the focus of much ongoing research
interest. Recent progress in inference has made ARG-based approaches feasible
across of range of applications, and many new methods using inferred ARGs as
input have appeared. This progress on the long-standing problem of ARG
inference has proceeded in two distinct directions. First,
the Bayesian inference of ARGs under the Sequentially Markov
Coalescent (SMC), is now practical for tens-to-hundreds of samples.
Second, approximate models and heuristics can now scale to sample sizes two to three
orders of magnitude larger. Although these heuristic methods are reasonably
accurate under many metrics, one significant drawback is that the ARGs they
estimate do not have the topological properties required to compute a
likelihood under models such as the SMC under present-day formulations.
In particular, heuristic inference methods typically do not estimate
precise details about recombination events, which are currently
required to compute a likelihood.
In this paper we present a
backwards-time formulation of the SMC (conventionally regarded as
an along-the-genome process)
and derive a straightforward definition
of the likelihood of a general class of ARG under this model.
We show that this formulation does not require precise details of recombination events
to be estimated, and is robust to the presence of polytomies.
We discuss the possibilities
for ARG inference that this new formulation opens. \revpoint{2}{3}
\end{abstract}
